General purpose gene annotation using deep learning

### 2020-04-21 paper/main.tex
switch to bibtex

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}

@@

\section{Introduction}
For many tasks, it is necessary to have DNA regions labeled according to their function. This process is called gene annotation and it can be performed in different levels of precision, from simple coding -- non-coding classification to detailed structural labeling, such as enhancer regions or long non-coding RNA sequences in eukaryotes. In this work on eucaryotic genomes, we differentiate between Intergenic regions, Untranslated regions (UTRS, parts at the beginning and end of gene, that are transcribed but not translated), Coding sequences (CDS) and Introns. Because of the sheer size of genomes alone, gene annotation is a very labour intensive process and genomic database providers like NCBI [\citealp{NCBI13}] and Ensembl [\citealp{Ensembl16}] offer gene annotation pipelines. These pipelines integrate experimental data (from e.g. RNA-seq or proteogenomics) with homolog sequences in the database and ab-initio gene predictions. The latter is an attractive approach, because it is cheap and fast. State of the art performance ab-initio gene prediction is achieved by higher order hidden markov models (HMMs), such as Genscan [\citealp{Genscan}] or Augustus [\citealp{Augustus}] but their accuracy leaves room for improvement. By choosing possible states and transitions, designers of HMMs assume structure in the sequence that may limit its predictive power. Indeed, in practice HMMs have problems with generalizing across species and it seems actual learning of underlying patterns is limited. In the last decade, deep neural networks have been applied with tremendous success in many areas of statistical modelling. For sequence data, such as DNA, recurrent neural networks (RNN) in the form of long-short term memory (LSTM) are an established architecture and here, we employ them to improve on HMMs for ab-initio genome annotation.

RNNs have already shown promising results in related, but generally more limited investigations. [\citealp{Choudhary17}] carries out preliminary explorations on the potential of BLSTMs for cross species gene prediction, but trains his model on human genes only and tests it on two more species. DeepAnnotator [\citealp{DeepAnnotator}] uses bidirectional LSTMs (BLSTM) for gene finding in prokaryotes. Gene prediction in prokaryotes is considered more amenable than in eukaryotes, as genes in prokaryotes are proportionately more frequent in the genome, feature simpler control structures and do not use splicing [\citealp{Zhuo04}]. DanQ [\citealp{DanQ}] proposes the use of a BLSTM after a concurrent neural net (CNN) to find detailed motifs in the human genome. DeePromotor[\citealp{DeePromotor}] trains a similar architecture for the recognition of promotor regions. Recently, several groups [\citealp{SpliceAi}, \citealp{SpliceFinder}] successfully used CNNs to find splicing sites.

@@

\begin{table}[!t]
\processtable{Data group statistics for all data\label{tab:statistics_all}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
Average genome size in Gbp& 2.936 (+- 1.562) & 0.787 (+- 0.995) \\
Average gene length & 31,223 (+- 13,974)& 3,368 (+- 1,510)\\
Geenuff error rate & 0.311 (+- 0.129) & 0.351 (+- 0.249) \\
Fraction of class Intergenic  & 0.777 (+- 0.042) & 0.799 (+- 0.089) \\
Fraction of class UTR & 0.006 (+- 0.006) & 0.017 (+- 0.016) \\
Fraction of class CDS & 0.016 (+- 0.014) & 0.085 (+- 0.071) \\
Fraction of class Intron & 0.201 (+- 0.035) & 0.099 (+- 0.057) \\
\botrule
\end{tabular}}{Values are averages of the individual values of each genome in a data group. All statistics except the average gene lengths are excluding FASTA sequences without a gene and subsequences that were masked as completely erroneous. Each strand of DNA was counted seperately. The average gene length was determined by the average length of the longest transcript by cumulative exon length, which is how we query for genes during data generation. Brackets show the standart deviation.}
\end{table}

@@

We use two metrics to judge model performance. The {\it Genic F1} is our primary metric that provides the most comprehensive picture of annotation quality in one number. The {\it Subgenic F1} is similar to the Genic F1, except that it does not take UTR predictions into account. This is done for comparability with AUGUSTUS and further explained in Section \ref{sec:augustus}. Both metrics work by summing up True Positives (TP), False Positives (FP) and False Negatives (FN) of each considered class before calculating precision and recall from these values and combining the two into a F1 score. Both metrics do not take the values of the intergenic class into account, as this class is very abundant and appears to be by far the easiest to predict. (See Table \ref{tab:statistics_all} for the class distribution and Table \ref{tab:detailed_results} for a more in-depth report our model performance including the intergenic class). In the case of the Genic F1, this means that effectively only the TP of the intergenic class are disregarded. The metrics essentially provide a weighted mean of the performances in the considered classes.

\subsection{Model Architecture and Usage}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer normalization (citation ? ) between each BLSTM layer to produce the predictions in the form of a base pair wise classification. The model consists of circa 5.4 million parameters and was implemented with the deep learning library keras (citation) on top of TensorFlow (citation). We tested multiple different architectures before arriving at this model configuration, including convolutional neural networks (CNN) and hybrid architectures. The hyperparameters were determined by a combination of manual and automatic hyperparameter optimization (talk about TPE?, quote nni?). All relevant hyperparameters can be found in Table \ref{tab:params} or in the Helixer source code repository.

@@

\begin{table}[!t]
\processtable{Data arrays generated and used by Helixer\label{tab:arrays}} {
\begin{tabular}{@{}ll@{}}
\toprule Name & Information \\
\midrule
Input & Genomic sequence in the 4-dimensional IUPAC encoding\\
	  & (one hot encoding for non-ambiguous bases) \\
Output & Labels in a 4-dimensional one hot encoding representing \\
      & the classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}\\
Sample weights & One of $\{0,1\}$; whether there is an error at a base\\
\botrule
\end{tabular}}{An intron inside a UTR region is encoded as intron and can be inferred from the prediction order.}
\end{table}

@@

\begin{figure}[!tpb]
	\centerline{\includegraphics[width=0.45\textwidth]{images/overall_boxplot}}
	\caption{Subgenic F1 scores for all animal and plant genomes that were not used during training of Helixer, which is equivalent to all validation and test genomes. Results are given for Helixer and AUGUSTUS with the median predictions scores being denoted by the red lines. Details of how AUGUSTUS was used are given in Section \ref{sec:augustus}.}\label{fig:boxplot}
\end{figure}

\begin{table}[!t]
\processtable{Experimental Results and Comparison\label{tab:results}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
AUGUSTUS & 0.632 (+-0.09) & 0.757 (+-0.182)\\
Dilated CNN Baseline & - (+-) & - (+-)\\
Best Helixer model & - (+-) & - (+-) \\
Best model + varied input length & - (+-) & - \\
Best model (+ varied input length) + overlapping &  0.85 (+-0.068) & 0.843 (+-0.151) \\
(Varied input length +) overlapping + ensembling & 0.868 (+-0.064) & 0.863 (+-0.148) \\
\botrule
\end{tabular}}{Values are the median in Subgenic F1 across all validation and test genomes of the respective group. Varied input length was only used in the animal case. The best model was chosen out of the eight models of the ensemble for having the best performance in the reported setup.}
\end{table}

@@

\begin{thebibliography}{}

\bibitem[Thibaud-Nissen {\it et~al}, 2013]{NCBI13}
Thibaud-Nissen,F., Souvorov,A., Murphy,T., DiCuccio,M., Kitts,P. (2013) Eukaryotic Genome Annotation Pipeline{\it The NCBI Handbook, 2nd}

\bibitem[Aken {\it et~al}, 2016]{Ensembl16}
Aken,B., Ayling,S., Barrell,D., Clarke,L., {\it et~al.} (2016) The Ensembl gene annotation system, {\it Database}, {\bf 2016}, baw093

\bibitem[Burge {\it et} Karlin, 1997]{Genscan}
Burge,C., Karlin,S. (1997) Prediction of complete gene structures in human genomic DNA, {\it Journal of Molecular Biology}, {\bf 268-1}, 78-94

\bibitem[Stanke {\it et} Waack, 2003]{Augustus}
	Stanke,M., Waack,S., (2003) Gene prediction with a hidden Markov model and a new intron submodel, {\it Bioinformatics}, {\bf 19 Suppl.2}, ii215–ii225

\bibitem[Zhuo {\it et~al}., 2004]{Zhuo04}
Zhuo,W., Yazhu,C., Yixue,L. (2004) A Brief Review of Computational Gene Prediction Methods, {\it Genomics, Proteomics \& Bioinformatics}, 216-221

\bibitem[Amin {\it et~al}., 2018]{DeepAnnotator}
Amin,M., Yurovsky,A., Tian,Y., Skiena,S. (2018) DeepAnnotator: Genome Annotation with Deep Learning, {\it Proceedings of the 2018 ACM International Conference on Bioinformatics, Computational Biology, and Health Informatics}, 254-259.

\bibitem[Choudhary, 2017]{Choudhary17}
Choudhary,S. (2017) Predicting protein coding boundaries using RNNs, http://www.saket-choudhary.me/rnn-cds-prediction/predicting-protein-coding.pdf (accessed 2020-04-11)

\bibitem[Daniel {\it et~al}., 2016]{DanQ}
Daniel,Q., Xiaohui,X., (2016) DanQ: a hybrid convolutional and recurrent deep neural network for quantifying the function of DNA sequences, {\it Nucleic Acids Research}, {\bf 44-11}, e107

\bibitem[Oubounyt {\it et~al}., 2019]{DeePromotor}
Oubounyt,M., Louadi,Z., Tayara,H., Chong2,K. (2019) DeePromoter: Robust Promoter Predictor Using Deep Learning., {\it Frontiers in genetics}, {\bf 10}, 286

\bibitem[Jaganathan {\it et~al}., 2019]{SpliceAi}
Jaganathan,K., Panagiotopoulou,S., McRae,J., Darbandi,S., et.al., Predicting Splicing from Primary Sequence with Deep Learning, {\it Cell}, {\bf 176-3}, 535-548.e24

\bibitem[Wang {\it et~al}., 2019]{SpliceFinder}
Wang,R., Wang,Z., Wang,J. (2019) SpliceFinder: ab initio prediction of splice sites using convolutional neural network, {\it BMC Bioinformatics}, {\bf 20}, 652

\bibitem[Gupta {\it et~al}., 2017]{dCNNGupta}
Gupta,A., Rush.,A. (2017) Dilated convolutions for modeling long-distance genomic dependencies, {\it arXiv preprint arXiv:1710.01278}

\bibitem[Denton {\it et~al}., 2019]{Denton19}
Denton,A., Stiehler,F. (2019) Article title, {\it Journal Name}, {\bf 199}, 133-154.

\bibitem[Howe {\it et~al}., 2019]{Howe19}
Howe KL, Contreras-Moreira B, De Silva N, Maslen G, Akanni W, Allen J, Alvarez-Jarreta J, Barba M, Bolser DM, Cambell L, Carbajo M, Chakiachvili M, Christensen M, Cummins C, Cuzick A, Davis P, Fexova S, Gall A, George N, Gil L, Gupta P, Hammond-Kosack KE, Haskell E, Hunt SE, Jaiswal P, Janacek SH, Kersey PJ, Langridge N, Maheswari U, Maurel T, McDowall MD, Moore B, Muffato M, Naamati G, Naithani S, Olson A, Papatheodorou I, Patricio M, Paulini M, Pedro H, Perry E, Preece J, Rosello M, Russell M, Sitnik V, Staines DM, Stein J, Tello-Ruiz MK, Trevanion SJ, Urban M, Wei S, Ware D, Williams G, Yates AD, Flicek P. (2019) Ensembl Genomes 2020-enabling non-vertebrate genomic research. {\it Nucleic Acids Research 2019} https://doi.org/10.1093/nar/gkz890

\bibitem[Goodstein {\it et~al}., 2012]{Phytozome}
	David M. Goodstein, Shengqiang Shu, Russell Howson, Rochak Neupane, Richard D. Hayes, Joni Fazo, Therese Mitros, William Dirks, Uffe Hellsten, Nicholas Putnam, Daniel S. Rokhsar (2012) Phytozome: a comparative platform for green plant genomics, {\it Nucleic Acids Research}, Volume 40, Issue D1, Pages D1178–D1186, https://doi.org/10.1093/nar/gkr944

\end{thebibliography}

@@

\botrule
\end{tabular}}{Above are all parameters used during neural architecture search for the dilated CNN baseline. The same overall space was used for plant and animal data. We did, however, run multiple distinct searches that sometimes only operated over a subset of the given parameters. This was done as those seemed to be the most promising. We for example restricted the search space of the number of LSTM layers to the highest 3 values in later runs. Decisions were guided by the Genic F1 on the validation set of exactly the same data we trained our final LSTM models with. Runs with a Genic F1 below 0.5 after 10 epochs were stopped and the overall maximum epoch was 15. The performances seemed to be leveling off well before that. The batch size used for almost all runs was 32, the dilation was capped to 81 and the size of each fully connected layer was fixed at 128. New parameters where chosen at random. The implementation and parameter space is very roughly based on [\citealp{dCNNGupta}] and can be found in the Helixer source code repository. In total, we trained 30 models for the animals and 48 with the plant data.}
\end{table}

\begin{table}[!t]
\processtable{Final Parameters of dilated CNN baselines\label{tab:dcnn_best}} {
\begin{tabular}{@{}lll@{}}
\toprule Parameter & Animals & Plants\\
\midrule
Kernel Size & 16 & 16\\
Initial Filter Depth & 96 & 32\\
Number of Layers Before Doubling Filter Count & 2 & 2\\
Dilation Multiplier & 3 & 3\\
Number of Convolutional Layers & 6 & 8\\
Number of Hidden Fully Connected (FC) Layers & 2 & 0\\
Dropout Used on FC Output & 0.0 & 0.0\\
Learning Rate & 1e-4 & 1e-3\\
\botrule
\end{tabular}}{The parameters of the dilated CNN model that performed the best according to the Genic F1 of the validation data of the respective training genomes. The best animal model has circa 4.6 million parameters; the best plant model around 2.1 million.}
\end{table}

@@

\begin{table}[!t]
\processtable{Hyperparameters\label{tab:params}} {
\begin{tabular}{@{}ll@{}}
\toprule Name & Value \\
\midrule
Layers & 4\\
Units per layer & 256\\
Learning rate & 1e-4\\
Optimizer & Adam\\
\botrule
\end{tabular}}{}
\end{table}

\begin{table}[!t]
\processtable{Class weights\label{tab:params_cw}} {
\begin{tabular}{@{}lll@{}}
\toprule Class & Animal Value & Plant Value \\
\midrule
Intergenic & 0.7 & 0.3 \\
UTR & 1.6 & 1.0 \\
CDS & 1.2 & 0.9 \\
Intron & 1.2 & 0.3 \\
\botrule
\end{tabular}}{The class weights are the only hyperparameters that are different for animals and plants. The exact weights were determined manually and partially reflect the difference in the class distributions of both datasets (see Table \ref{tab:statistics_all}). All other hyperparameters are shown in Table \ref{tab:params}.}
\end{table}

\subsection{Training Data statistics}
\label{sec:training_data}
\begin{table}[!t]
\processtable{Data group statistics for our training data\label{tab:statistics_training}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
Average genome size in Gbp& 2.489 (+- 2.073) & 0.914 (+- 0.934) \\
Average gene length & 25,672 (+- 16,605) & 3,509 (+- 906)\\
Geenuff error rate & 0.253 (+- 0.138) & 0.134 (+- 0.072) \\
Fraction of class Intergenic & 0.752 (+- 0.05) & 0.808 (+- 0.106) \\
Fraction of class UTR & 0.013 (+- 0.011) & 0.033 (+- 0.023) \\
Fraction of class CDS & 0.028 (+- 0.028) & 0.077 (+- 0.052) \\
Fraction of class Intron  & 0.207 (+- 0.023) & 0.083 (+- 0.037) \\
\botrule
\end{tabular}}{The description of Table \ref{tab:statistics_all} applies here as well.}
\end{table}

@@

\begin{table}[!t]
\processtable{Class weights\label{tab:detailed_results}} {
\begin{tabular}{@{}lll@{}}
\toprule Class & Animal Value & Plant Value \\
\midrule
Intergenic & 0.7 & 0.3 \\
UTR & 1.6 & 1.0 \\
CDS & 1.2 & 0.9 \\
Intron & 1.2 & 0.3 \\
\botrule
\end{tabular}}{}
\end{table}

>>>>

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}
\usepackage{url}

@@

\section{Introduction}
For many tasks, it is necessary to have DNA regions labeled according to their function. This process is called gene annotation and it can be performed in different levels of precision, from simple coding -- non-coding classification to detailed structural labeling, such as enhancer regions or long non-coding RNA sequences in eukaryotes. In this work on eucaryotic genomes, we differentiate between Intergenic regions, Untranslated regions (UTRS, parts at the beginning and end of gene, that are transcribed but not translated), Coding sequences (CDS) and Introns. Because of the sheer size of genomes alone, gene annotation is a very labour intensive process and genomic database providers like NCBI \citep{thibaud2013eukaryotic} and Ensembl \citep{aken2016ensembl} offer gene annotation pipelines. These pipelines integrate experimental data (from e.g. RNA-seq or proteogenomics) with homolog sequences in the database and ab-initio gene predictions. The latter is an attractive approach, because it is cheap and fast. State of the art performance ab-initio gene prediction is achieved by higher order hidden markov models (HMMs), such as Genscan \citep{burge1997prediction} or Augustus \citep{stanke2003gene} but their accuracy leaves room for improvement. By choosing possible states and transitions, designers of HMMs assume structure in the sequence that may limit its predictive power. Indeed, in practice HMMs have problems with generalizing across species and it seems actual learning of underlying patterns is limited. In the last decade, deep neural networks have been applied with tremendous success in many areas of statistical modelling. For sequence data, such as DNA, recurrent neural networks (RNN) in the form of long-short term memory (LSTM) are an established architecture and here, we employ them to improve on HMMs for ab-initio genome annotation.

RNNs have already shown promising results in related, but generally more limited investigations. \citep{choudharypredicting} carries out preliminary explorations on the potential of BLSTMs for cross species gene prediction, but trains his model on human genes only and tests it on two more species. DeepAnnotator \citep{amin2018deepannotator} uses bidirectional LSTMs (BLSTM) for gene finding in prokaryotes. Gene prediction in prokaryotes is considered more amenable than in eukaryotes, as genes in prokaryotes are proportionately more frequent in the genome, feature simpler control structures and do not use splicing \citep{wang2004brief}. DanQ \citep{quang2016danq} proposes the use of a BLSTM after a concurrent neural net (CNN) to find detailed motifs in the human genome. DeePromotor \citep{oubounyt2019deepromoter} trains a similar architecture for the recognition of promotor regions. Recently, several groups \citep{jaganathan2019predicting, wang2019splicefinder} successfully used CNNs to find splicing sites.

@@

\begin{table}[!t]
\processtable{Data group statistics for all data\label{tab:statistics_all}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
Average genome size in Gbp& 2.936 (+- 1.562) & 0.787 (+- 0.995) \\
Average gene length & 31,223 (+- 13,974)& 3,368 (+- 1,510)\\
Geenuff error rate & 0.311 (+- 0.129) & 0.351 (+- 0.249) \\
Fraction of class Intergenic  & 0.777 (+- 0.042) & 0.799 (+- 0.089) \\
Fraction of class UTR & 0.006 (+- 0.006) & 0.017 (+- 0.016) \\
Fraction of class CDS & 0.016 (+- 0.014) & 0.085 (+- 0.071) \\
Fraction of class Intron & 0.201 (+- 0.035) & 0.099 (+- 0.057) \\
\botrule
\end{tabular}}{{\it Note:} Values are averages of the individual values of each genome in a data group. All statistics except the average gene lengths are excluding FASTA sequences without a gene and subsequences that were masked as completely erroneous. Each strand of DNA was counted seperately. The average gene length was determined by the average length of the longest transcript by cumulative exon length, which is how we query for genes during data generation. Brackets show the standart deviation.}
\end{table}

@@

\subsection{Metrics}
We use two metrics to judge model performance. The {\it Genic F1} is our primary metric that provides the most comprehensive picture of annotation quality in one number. The {\it Subgenic F1} is similar to the Genic F1, except that it does not take UTR predictions into account. This is done for comparability with AUGUSTUS and further explained in Section \ref{sec:augustus}. Both metrics work by summing up True Positives (TP), False Positives (FP) and False Negatives (FN) of each considered class before calculating precision and recall from these values and combining the two into a F1 score. Both metrics do not take the values of the intergenic class into account, as this class is very abundant and appears to be by far the easiest to predict. (See Table \ref{tab:statistics_all} for the class distribution and Table \ref{tab:detailed_results} for a more in-depth report our model performance including the intergenic class). In the case of the Genic F1, this means that effectively only the TP of the intergenic class are disregarded. The metrics essentially provide a weighted mean of the performances in the considered classes.

\subsection{Model Architecture and Usage}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer normalization (\citealp{Layernorm}) between each BLSTM layer to produce the predictions in the form of a base pair wise classification. The model consists of circa 5.4 million parameters and was implemented with the deep learning library keras (citation) on top of TensorFlow (\citealp{Tensorflow}). We tested multiple different architectures before arriving at this model configuration, including convolutional neural networks (CNN) and hybrid architectures. The hyperparameters were determined by a combination of manual and automatic hyperparameter optimization (talk about TPE?, quote nni?). All relevant hyperparameters can be found in Table \ref{tab:params} or in the Helixer source code repository.

@@

\begin{table}[!t]
\processtable{Data arrays generated and used by Helixer\label{tab:arrays}} {
\begin{tabular}{@{}ll@{}}
\toprule Name & Information \\
\midrule
Input & Genomic sequence in the 4-dimensional IUPAC encoding\\
	  & (one hot encoding for non-ambiguous bases) \\
Output & Labels in a 4-dimensional one hot encoding representing \\
      & the classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}\\
Sample weights & One of $\{0,1\}$; whether there is an error at a base\\
\botrule
\end{tabular}}{{\it Note:} An intron inside a UTR region is encoded as intron and can be inferred from the prediction order.}
\end{table}

@@

\begin{figure}[!tpb]
	\centerline{\includegraphics[width=0.45\textwidth]{images/overall_boxplot}}
	\caption{Boxplots comparing the Subgenic F1 scores for all animal and plant genomes that were not used during training of Helixer, which is equivalent to all validation and test genomes. Results are given for Helixer and AUGUSTUS with the median predictions scores being denoted by the red lines. Details of how AUGUSTUS was used are given in Section \ref{sec:augustus}.}\label{fig:boxplot}
\end{figure}

\begin{table}[!t]
\processtable{Experimental Results and Comparison\label{tab:results}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
AUGUSTUS & 0.632 (+-0.09) & 0.757 (+-0.182)\\
Dilated CNN Baseline & - (+-) & - (+-)\\
Best Helixer model & - (+-) & - (+-) \\
Best model + varied input length & - (+-) & - \\
Best model (+ varied input length) + overlapping &  0.85 (+-0.068) & 0.843 (+-0.151) \\
(Varied input length +) overlapping + ensembling & 0.868 (+-0.064) & 0.863 (+-0.148) \\
\botrule
\end{tabular}}{{\it Note:} Values are the median in Subgenic F1 across all validation and test genomes of the respective group. Varied input length was only used in the animal case. The best model was chosen out of the eight models of the ensemble for having the best performance in the reported setup.}
\end{table}

@@

\bibliographystyle{natbib}
\bibliography{literature}

@@

\botrule
\end{tabular}}{{\it Note:} Above are all parameters used during neural architecture search for the dilated CNN baseline. The same overall space was used for plant and animal data. We did, however, run multiple distinct searches that sometimes only operated over a subset of the given parameters. This was done as those seemed to be the most promising. We for example restricted the search space of the number of LSTM layers to the highest 3 values in later runs. Decisions were guided by the Genic F1 on the validation set of exactly the same data we trained our final LSTM models with. Runs with a Genic F1 below 0.5 after 10 epochs were stopped and the overall maximum epoch was 15. The performances seemed to be leveling off well before that. The batch size used for almost all runs was 32, the dilation was capped to 81 and the size of each fully connected layer was fixed at 128. New parameters where chosen at random. The implementation and parameter space is very roughly based on [\citealp{dCNNGupta}] and can be found in the Helixer source code repository. In total, we trained 30 models for the animals and 48 with the plant data.}
\end{table}

\begin{table}[!t]
\processtable{Final Parameters of dilated CNN baselines\label{tab:dcnn_best}} {
\begin{tabular}{@{}lll@{}}
\toprule Parameter & Animals & Plants\\
\midrule
Kernel Size & 16 & 16\\
Initial Filter Depth & 96 & 32\\
Number of Layers Before Doubling Filter Count & 2 & 2\\
Dilation Multiplier & 3 & 3\\
Number of Convolutional Layers & 6 & 8\\
Number of Hidden Fully Connected (FC) Layers & 2 & 0\\
Dropout Used on FC Output & 0.0 & 0.0\\
Learning Rate & 1e-4 & 1e-3\\
\botrule
\end{tabular}}{{\it Note:} The parameters of the dilated CNN model that performed the best according to the Genic F1 of the validation data of the respective training genomes. The best animal model has circa 4.6 million parameters; the best plant model around 2.1 million.}
\end{table}

@@

\begin{table}[!t]
\processtable{Hyperparameters\label{tab:params}} {
\begin{tabular}{@{}ll@{}}
\toprule Name & Value \\
\midrule
Layers & 4\\
Units per layer & 256\\
Learning rate & 1e-4\\
Optimizer & Adam\\
\botrule
\end{tabular}}{{\it Note:} }
\end{table}

\begin{table}[!t]
\processtable{Class weights\label{tab:params_cw}} {
\begin{tabular}{@{}lll@{}}
\toprule Class & Animal Value & Plant Value \\
\midrule
Intergenic & 0.7 & 0.3 \\
UTR & 1.6 & 1.0 \\
CDS & 1.2 & 0.9 \\
Intron & 1.2 & 0.3 \\
\botrule
\end{tabular}}{{\it Note:} The class weights are the only hyperparameters that are different for animals and plants. The exact weights were determined manually and partially reflect the difference in the class distributions of both datasets (see Table \ref{tab:statistics_all}). All other hyperparameters are shown in Table \ref{tab:params}.}
\end{table}

\subsection{Training Data statistics}
\label{sec:training_data}
\begin{table}[!t]
\processtable{Data group statistics for our training data\label{tab:statistics_training}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
Average genome size in Gbp& 2.489 (+- 2.073) & 0.914 (+- 0.934) \\
Average gene length & 25,672 (+- 16,605) & 3,509 (+- 906)\\
Geenuff error rate & 0.253 (+- 0.138) & 0.134 (+- 0.072) \\
Fraction of class Intergenic & 0.752 (+- 0.05) & 0.808 (+- 0.106) \\
Fraction of class UTR & 0.013 (+- 0.011) & 0.033 (+- 0.023) \\
Fraction of class CDS & 0.028 (+- 0.028) & 0.077 (+- 0.052) \\
Fraction of class Intron  & 0.207 (+- 0.023) & 0.083 (+- 0.037) \\
\botrule
\end{tabular}}{{\it Note:} The description of Table \ref{tab:statistics_all} applies here as well.}
\end{table}

@@

\begin{table}[!t]
\processtable{Class weights\label{tab:detailed_results}} {
\begin{tabular}{@{}lll@{}}
\toprule Class & Animal Value & Plant Value \\
\midrule
Intergenic & 0.7 & 0.3 \\
UTR & 1.6 & 1.0 \\
CDS & 1.2 & 0.9 \\
Intron & 1.2 & 0.3 \\
\botrule
\end{tabular}}{{\it Note:} }
\end{table}
